# Appendix: Introduction to JAGS and BUGS language

## Why learn JAGS?

So far, we have written Monte Carlo and Gibbs samplers directly in R. This is useful for understanding the mechanics of Bayesian computation, but it *becomes inconvenient when models are more complicated*.

**JAGS** stands for **Just Another Gibbs Sampler**. It is a program for fitting Bayesian hierarchical models using **MCMC**, especially Gibbs sampling and related methods.

JAGS uses the **BUGS language**, where BUGS stands for **B**ayesian inference **U**sing **G**ibbs **S**ampling.

The basic idea is simple:

1.  we describe the statistical model in BUGS syntax;
2.  we provide the observed data;
3.  we choose which parameters to monitor;
4.  JAGS runs the MCMC algorithm and returns posterior samples.


## What BUGS language does

BUGS language is a **model specification language**. It is not a general programming language like R or Python.

Its role is to describe:

-   the **sampling model**
-   the **prior distributions**
-   deterministic relationships among parameters
-   optional **derived quantities**

For example:

$$
Y_i \mid \theta \sim \text{Poisson}(\theta), \qquad
\theta \sim \text{Gamma}(a,b)
$$

Instead of writing the Gibbs sampler ourselves, we simply describe the model and JAGS handles the sampling.

## History

-   BUGS was developed in the 1990s by David Spiegelhalter and colleagues at the University of Cambridge.

-   JAGS was created by Martyn Plummer in the early 2000s as a more flexible and open-source alternative to BUGS.

-  Both BUGS and JAGS have been widely used in academia and industry for Bayesian data analysis, especially in fields like ecology, epidemiology, and social sciences.

-  JAGS is designed to be compatible with BUGS models, so most BUGS code can be run in JAGS with little or no modification.

-  JAGS is often used in conjunction with R through packages like `rjags` and `jagsUI`, which provide an interface for running JAGS models from R.

-  The BUGS language has influenced the development of other probabilistic programming languages, such as **Stan**, **nimble** and **PyMC3**, which also allow users to specify Bayesian models in a high-level syntax.

- JAGS continues to be maintained and updated, with the latest version being 4.3.2 as of 2026.

- `coda` package is often used to analyze MCMC output from JAGS, providing tools for convergence diagnostics and posterior summaries.

## Basic structure of a JAGS model

A JAGS model is written inside a `model {}` block.

``` r
model {
  for (i in 1:n) {
    y[i] ~ dpois(theta)
  }

  theta ~ dgamma(a, b)
}
```

The first part describes the likelihood, and the second part specifies
the prior.



## BUGS syntax rules

**Stochastic nodes**

Random variables are written using `~`.

``` r
theta ~ dgamma(a, b)
y[i] ~ dpois(theta)
```

**Deterministic nodes**

Deterministic relationships use `<-`.

``` r
sigma <- 1 / sqrt(tau)
mu[i] <- alpha + beta * x[i]
```

**Loops**

Repeated data are handled with loops.

``` r
for (i in 1:n) {
  y[i] ~ dnorm(mu, tau)
}
```

**Comments**

Use `#` for comments.

``` r
# prior distribution
theta ~ dnorm(0, 0.01)
```

::: {.callout-warning title="JAGS syntax"}

Pay attention to the JAGS syntax. Sometimes it may be different from what you expect! For instance, JAGS uses **precision** rather than variance for normal distributions `dnorm(mean, precision)`. Precision is

$$
\tau = \frac{1}{\sigma^2}.
$$

or in the lecture, we used $\tilde{\sigma}^2=1/\sigma^2$.

In R code, we write

``` r
tau <- 1 / sigma^2
```

:::


Table: Common probability distributions in JAGS (BUGS language)

| Distribution | JAGS syntax |
|--------------|-------------|
| Normal       | `dnorm(mu, tau)` |
| Poisson      | `dpois(lambda)` |
| Binomial     | `dbin(p, n)` |
| Bernoulli    | `dbern(p)` |
| Gamma        | `dgamma(a, b)` |
| Beta         | `dbeta(a, b)` |
| Uniform      | `dunif(a, b)` |


## Example: Beta--Binomial model

$$
Y \mid \theta \sim \text{Binomial}(n,\theta),
\qquad
\theta \sim \text{Beta}(a,b)
$$

JAGS model:

``` r
model {
  y ~ dbin(theta, n)
  theta ~ dbeta(a, b)
}
```

For multiple observations:

``` r
model {
  for (i in 1:n) {
    y[i] ~ dbern(theta)
  }
  theta ~ dbeta(a, b)
}
```


::: {.callout-example title="Poisson--Gamma model"}


$$
Y_i \mid \theta \sim \text{Poisson}(\theta),
\qquad
\theta \sim \text{Gamma}(a,b)
$$

JAGS:

``` r
model {
  for (i in 1:n) {
    y[i] ~ dpois(theta)
  }

  theta ~ dgamma(a, b)
}
```

:::


::: {.callout-example title="Normal model"}


$$
Y_i \mid \theta,\tau \sim \text{Normal}(\theta,\tau)
$$

Priors

$$
\theta \sim \text{Normal}(\mu_0,\tau_0), \qquad
\tau \sim \text{Gamma}(a,b)
$$

JAGS model:

``` r
model {
  for (i in 1:n) {
    y[i] ~ dnorm(theta, tau)
  }

  theta ~ dnorm(mu0, tau0)
  tau ~ dgamma(a, b)

  sigma <- 1 / sqrt(tau)
}
```

:::


## Data list in R

``` r
data_jags <- list(
  y = y,
  n = length(y),
  a = 2,
  b = 1
)
```



## Parameters to monitor

``` r
params <- c("theta")
```



## Example JAGS workflow

``` r
library(jagsUI)

fit <- jags(
  data = data_jags,
  parameters.to.save = params,
  model.file = textConnection(model_string),
  n.chains = 3,
  n.iter = 5000,
  n.burnin = 1000
)

print(fit)
```


**Initial values**

Optional but sometimes useful.

``` r
inits <- function() {
  list(theta = 1)
}
```

**Posterior output**

Typical outputs include

-   posterior means
-   credible intervals
-   trace plots
-   density plots
-   effective sample size
-   $\hat{R}$ convergence diagnostic



Table: Useful functions in the BUGS/JAGS language

| Function | Meaning |
|---------|---------|
| `log(x)` | logarithm |
| `exp(x)` | exponential |
| `sqrt(x)` | square root |
| `pow(x, a)` | power |
| `step(x)` | indicator function |

Example:

``` r
tau <- pow(sigma, -2)
```



## Indexing

Sometimes, for each of the data $y_i$, you may want to have a different prior mean $\mu_i$ that depends on covariates $x_i$. This can be done as follows

``` r
for (i in 1:n) {
  y[i] ~ dnorm(mu[i], tau)
  mu[i] <- alpha + beta * x[i]
}
```


::: {.callout-example title="A completely worked example in JAGS: Poisson--Gamma model"}

In this section we fit a simple Bayesian Poisson model in **JAGS**.
Suppose

$$
Y_1,\ldots,Y_n \mid \theta \sim \text{i.i.d. Poisson}(\theta),
$$

with prior

$$
\theta \sim \text{Gamma}(a,b),
$$

where the Gamma distribution is parameterized by **shape** $a$ and **rate** $b$.

We will:

1. specify the model in BUGS/JAGS language;
2. prepare the data in R;
3. run JAGS using `jagsUI`;
4. extract posterior samples;
5. compare the MCMC output with the exact posterior.


### Data

We use the following toy dataset:

```{r}
y <- c(3, 2, 4, 1, 3)
n <- length(y)

a <- 2
b <- 1
sum_y <- sum(y)

y
sum_y
```

Under the Poisson–Gamma model, the posterior distribution is available analytically:

$$
\theta \mid y \sim \text{Gamma}!\left(a+\sum_{i=1}^n y_i,; b+n\right).
$$

So in this example,

$$
\theta \mid y \sim \text{Gamma}(2+13,;1+5)=\text{Gamma}(15,6).
$$

This makes the example useful because we can compare the JAGS output to the exact answer.

**Step 1: write the JAGS model**


In BUGS language, the model is

``` {r}
model_string <- "
model {

  # likelihood
  for (i in 1:n) {
    y[i] ~ dpois(theta)
  }

  # prior
  theta ~ dgamma(a, b)

}
"
cat(model_string)
```

A few notes:

+ `dpois(theta)` means Poisson with mean $\theta$;
+ `dgamma(a, b)` in JAGS uses shape and rate;
+ `theta` is the parameter we want to estimate.


**Step 2: prepare the data and monitored parameters**

JAGS needs the data in a named list


``` {r}
data_jags <- list(
  y = y,
  n = n,
  a = a,
  b = b
)

params <- c("theta")
```

We will monitor only theta.

**Step 3: choose initial values**

Initial values are optional here, but we include them for completeness.

``` {r}
inits <- function() {
  list(theta = 2)
}
```

**Step 4: run JAGS**

``` {r}

library(jagsUI)

set.seed(8310)

fit <- jags(
  data = data_jags,
  inits = inits,
  parameters.to.save = params,
  model.file = textConnection(model_string),
  n.chains = 3,
  n.iter = 5000,
  n.burnin = 1000,
  n.thin = 2
)
```

Now print the summary:

``` {r}
print(fit)
```

This output includes posterior means, standard deviations, credible intervals, and convergence diagnostics.


**Step 5: compare with the exact posterior**

The exact posterior is

$$
\theta \mid y \sim \text{Gamma}(15,6).
$$

So the exact posterior mean is

$$
E(\theta \mid y) = \frac{15}{6} = 2.5.
$$

The exact posterior variance is

$$
\mathrm{Var}(\theta \mid y) = \frac{15}{6^2} = \frac{15}{36}.
$$

We can compute the exact summaries in R:

``` {r}
shape_post <- a + sum_y
rate_post  <- b + n

exact_mean <- shape_post / rate_post
exact_var  <- shape_post / rate_post^2
exact_ci   <- qgamma(c(0.025, 0.975), shape = shape_post, rate = rate_post)

exact_mean
exact_var
exact_ci
```

The JAGS summary for theta should be very close to these values.


**Step 6: extract posterior samples**

The posterior samples are stored inside the fitted object.

``` {r}
theta_post <- fit$samples[, , "theta"]
theta_post <- as.vector(theta_post)

head(theta_post)
length(theta_post)
```

These are MCMC draws from the posterior distribution of $\theta$.


**Step 7: posterior histogram and exact density**

Now compare the JAGS samples to the exact Gamma posterior density.

``` {r}
library(ggplot2)
theta_post <- as.numeric(as.matrix(fit$samples)[,"theta"])

theta_df <- data.frame(theta = theta_post)
ggplot(theta_df, aes(x = theta)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "grey80",
                 color = "white") +
  stat_function(
    fun = dgamma,
    args = list(shape = shape_post, rate = rate_post),
    linewidth = 1.2,
    color = "black"
  ) +
  labs(
    title = "Posterior of theta: JAGS samples vs exact density",
    x = expression(theta),
    y = "Density"
  ) +
  theme_minimal()
``` 

The histogram of the JAGS samples should align closely with the exact posterior density.


**Step 8: posterior probability of an event**

Suppose we want to estimate

$$
\Pr(\theta < 2.8 \mid y).
$$

From the JAGS samples:

``` {r}
mean(theta_post < 2.8)
```

From the exact posterior:

```{r}
pgamma(2.8, shape = shape_post, rate = rate_post)
```

These two values should be close.


**Step 9: posterior predictive simulation**

Suppose $\tilde Y$ is a future observation from the same population.
The posterior predictive distribution is

$$
p(\tilde y \mid y)
＝
\int p(\tilde y \mid \theta),p(\theta \mid y),d\theta.
$$

Using the JAGS samples, we can simulate from the posterior predictive distribution:

``` {r}
set.seed(8310)
y_tilde <- rpois(length(theta_post), lambda = theta_post)

head(y_tilde)
mean(y_tilde)
var(y_tilde)


hist(y_tilde,
     breaks = seq(-0.5, max(y_tilde) + 0.5, by = 1),
     col = "grey85",
     border = "white",
     main = "Posterior predictive distribution",
     xlab = expression(tilde(Y)))
```
     
     
This histogram represents the distribution of a future observation $\tilde Y$ given the observed data $y$.



Interpretation

This example illustrates the full JAGS workflow:
	•	specify a Bayesian model in BUGS language;
	•	provide the data and priors;
	•	run MCMC in JAGS;
	•	summarize the posterior;
	•	compare with an exact answer when available.

Because the Poisson–Gamma model is conjugate, this example is especially useful for learning.
The JAGS output should agree closely with the exact posterior

$$
\text{Gamma}(15,6).
$$


**Full code in one chunk**

```{r full}
library(jagsUI)
library(coda)
library(ggplot2)

# ===============================
# Data
# ===============================
y <- c(3, 2, 4, 1, 3)
n <- length(y)
a <- 2
b <- 1
sum_y <- sum(y)

# ===============================
# JAGS model
# ===============================
model_string <- "
model {

  for (i in 1:n) {
    y[i] ~ dpois(theta)
  }

  theta ~ dgamma(a, b)

}
"

# ===============================
# Data list for JAGS
# ===============================
data_jags <- list(
  y = y,
  n = n,
  a = a,
  b = b
)

params <- c("theta")

inits <- function() {
  list(theta = 2)
}

# ===============================
# Run JAGS
# ===============================
set.seed(8310)

fit <- jags(
  data = data_jags,
  inits = inits,
  parameters.to.save = params,
  model.file = textConnection(model_string),
  n.chains = 3,
  n.iter = 5000,
  n.burnin = 1000,
  n.thin = 2
)

print(fit)

# ===============================
# Exact posterior
# ===============================
shape_post <- a + sum_y
rate_post  <- b + n

exact_mean <- shape_post / rate_post
exact_var  <- shape_post / rate_post^2
exact_ci   <- qgamma(c(0.025,0.975), shape = shape_post, rate = rate_post)

exact_mean
exact_var
exact_ci

# ===============================
# Extract posterior samples
# ===============================
theta_post <- as.numeric(as.matrix(fit$samples)[,"theta"])

theta_df <- data.frame(theta = theta_post)

# ===============================
# Posterior distribution plot
# ===============================
ggplot(theta_df, aes(x = theta)) +
  geom_histogram(aes(y = after_stat(density)),
                 bins = 30,
                 fill = "grey80",
                 color = "white") +
  stat_function(
    fun = dgamma,
    args = list(shape = shape_post, rate = rate_post),
    linewidth = 1.2,
    color = "black"
  ) +
  labs(
    title = "Posterior of theta: JAGS samples vs exact density",
    x = expression(theta),
    y = "Density"
  ) +
  theme_minimal()

# ===============================
# Posterior probability example
# ===============================
mean(theta_post < 2.8)

pgamma(2.8, shape = shape_post, rate = rate_post)

# ===============================
# Posterior predictive simulation
# ===============================
set.seed(8310)

y_tilde <- rpois(length(theta_post), lambda = theta_post)

pred_df <- data.frame(y_tilde = y_tilde)

# ===============================
# Posterior predictive plot
# ===============================
ggplot(pred_df, aes(x = y_tilde)) +
  geom_histogram(
    binwidth = 1,
    fill = "grey80",
    color = "white",
    boundary = -0.5
  ) +
  labs(
    title = "Posterior Predictive Distribution",
    x = expression(tilde(Y)),
    y = "Count"
  ) +
  theme_minimal()
```

:::




## Summary and Take home message

-   JAGS performs Bayesian inference using MCMC.
-   Models are written in the BUGS language.
-   `~` denotes a distribution.
-   `<-` denotes deterministic relationships.
-   Normal distributions use **precision** instead of variance.
-   JAGS separates **model**, **data**, and **parameters**.

::: {.callout-warning title="Common (beginner) mistakes"}

1.  Forgetting that normal uses **precision**
2.  Mixing deterministic `<-` and stochastic `~`
3.  Assuming all R syntax works inside JAGS
4.  Forgetting to pass constants in the data list

:::

---

Reference:

-   [JAGS User Manual](https://sourceforge.net/projects/mcmc-jags/files/JAGS/4.x/jags_user_manual.pdf)

-  [jagsUI package documentation](https://cran.r-project.org/web/packages/jagsUI/jagsUI.pdf)